\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

As Artificial Intelligence (AI) systems become increasingly integrated into critical societal domains, the need for robust, transparent, and multi-stakeholder ethical evaluation becomes paramount. This report presents the design and implementation of the Multi-stakeholder Ethical Decision Framework (MEDF), a software platform developed to address this challenge. MEDF provides a structured, reproducible, and auditable workflow for evaluating AI applications against prominent governance frameworks, including the EU AI Act Assessment List for Trustworthy AI (ALTAI), the NIST AI Risk Management Framework (RMF), and Singapore's Model AI Governance Framework (MGAF). The system operationalizes high-level ethical principles into quantifiable metrics across six unified dimensions: transparency and explainability, fairness and non-discrimination, safety and robustness, privacy and data governance, human agency and oversight, and accountability.

By enabling configurable stakeholder weightings, MEDF surfaces and quantifies disagreements in ethical priorities, employing multi-objective optimization techniques (NSGA-II) to identify Pareto-optimal consensus solutions. This work details the system's three-tier architecture, the algorithmic design of its scoring (TOPSIS, WSM) and conflict resolution engines, and its application to three real-world case studies: live facial recognition, automated hiring, and clinical diagnostics. The resulting platform serves as a decision-support tool for developers, regulators, and affected communities, fostering a more rigorous and inclusive approach to AI governance.

\textbf{Keywords:} Ethical AI, AI Governance, Decision Support Systems, Multi-Criteria Decision Making, TOPSIS, Stakeholder Analysis, Explainable AI, Algorithmic Fairness.
