\chapter{Verification Memo and Changelog}
\label{app:changelog}

This appendix documents the verification process applied to this report and the changes made relative to the initial draft. It is intended to provide an honest and transparent record of what was checked, what was corrected, and what limitations remain.

\section{Report Verification Process}

The following verification steps were performed to ensure the accuracy and consistency of this report:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Code-to-Report Verification.} The stakeholder weight table (Table~\ref{tab:weights}) was cross-referenced with the default values in \texttt{app/framework\_registry.py} and confirmed to match. The framework YAML dimension weights in Appendix~\ref{app:yaml} were verified against the actual \texttt{app/frameworks/*.yaml} files. The scoring algorithm descriptions in Chapter~\ref{ch:algorithms} were compared against the implementations in \texttt{app/scoring\_engine.py} and \texttt{app/routers/evaluate.py}. The conflict analysis descriptions were verified against \texttt{app/routers/conflicts.py}.

  \item \textbf{Score Reproduction.} All nine per-stakeholder, per-framework scores for each of the three case studies (27 scores total) were independently reproduced by calling the live MEDF API with the case study dimension scores and default stakeholder weights. The reproduced scores matched the values reported in Chapter~\ref{ch:case_studies} to four decimal places.

  \item \textbf{Conflict Matrix Reproduction.} The contribution-based Spearman $\rho$ values for all three case studies were independently reproduced via the \texttt{/api/conflicts} endpoint. The reproduced matrices matched the values reported in Chapter~\ref{ch:case_studies}.

  \item \textbf{Architectural Honesty Check.} The final repository was verified to implement conflict analysis directly in \texttt{app/routers/conflicts.py}. The report was revised to remove references to a standalone \texttt{app/conflict\_detection.py} module and to describe the router-based implementation truthfully.

  \item \textbf{Reference Audit.} All bibliographic references were verified for correctness. Author names, titles, publication venues, years, and DOIs were checked against the original sources. One incorrect citation was corrected (Floridi 2018).

  \item \textbf{Wording Normalization.} Terminology was standardized throughout the report. The project number was corrected from CCDS25-0323 to CCDS25-0582. Framework names were consistently rendered as \enquote{ALTAI}, \enquote{NIST AI RMF}, and \enquote{MGAF}.

  \item \textbf{Screenshot Verification.} All platform interface screenshots in Chapter~\ref{ch:interface} were captured from the live application running on localhost, using Playwright headless browser automation at 1920$\times$1080 resolution with 2$\times$ device pixel ratio.
\end{enumerate}

\section{Known Limitations of Verification}

The following items were \emph{not} independently verified and represent known limitations:

\begin{itemize}[leftmargin=2em]
  \item The baseline dimension scores for the three case studies were derived from qualitative analysis of public reports and are inherently subjective. No formal inter-rater reliability assessment was performed.
  \item The risk level labels returned by the API are \texttt{"N/A"} in the current implementation; the risk classification logic in the Streamlit frontend was not independently verified against the thresholds documented in Chapter~\ref{ch:algorithms}.
  \item The NSGA-II Pareto frontier results are stochastic (despite a fixed seed) and may vary slightly across different hardware or library versions.
\end{itemize}

\section{Changelog}

The following material changes were made during the revision process:

\begin{table}[H]
\centering
\caption{Revision changelog.}
\label{tab:changelog}
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
\textbf{Item} & \textbf{Change Description} \\
\midrule
Project Number & Corrected from CCDS25-0323 to CCDS25-0582 throughout. \\
\addlinespace
Academic Year & Corrected from 2024/2025 to 2025/2026 on the cover page. \\
\addlinespace
Ch7 Sec 7.3 & Corrected the conflict-analysis description to reflect the final repository state: \texttt{routers/conflicts.py} is the actual implementation, and no standalone \texttt{conflict\_detection.py} module is present in the final codebase. \\
\addlinespace
Ch7 Fig 7.2 & Simplified the module diagram to remove the non-existent \texttt{conflict\_detection.py} node and show the direct relationship between \texttt{routers/conflicts.py} and \texttt{harm\_assessment.py}. \\
\addlinespace
Ch8 Sec 8.5 & Split the single risk threshold table into two tables: one for the evaluation router (closeness coefficient) and one for the harm assessment module (harm score), reflecting the two different threshold systems in the code. \\
\addlinespace
Ch9 Sec 9.2.4 & Expanded the routers description to individually document \texttt{evaluate.py}, \texttt{conflicts.py}, and \texttt{pareto.py}, noting that \texttt{conflicts.py} is the de facto conflict engine. \\
\addlinespace
Ch11 Conflict Matrices & Replaced the identical weights-only matrices (which were the same across all three cases) with case-specific contribution-based matrices reproduced from the API. \\
\addlinespace
Ch11 Cross-Case Table & Updated overall scores and max conflict values to reflect contribution-based analysis. \\
\addlinespace
Ch4 Heading & Removed \enquote{Compulsory References:} prefix from section heading. \\
\addlinespace
Ch12 Heading & Changed \enquote{Key Findings} to \enquote{Critical Findings}. \\
\addlinespace
Appendix A & Corrected API example to use actual field names (\texttt{transparency\_explainability} instead of \texttt{transparency}) and actual request schema (\texttt{ai\_system} object). \\
\addlinespace
Appendix B & Corrected EU ALTAI YAML excerpt weights from (T=0.18, F=0.20, S=0.18, P=0.18, H=0.12, A=0.14) to actual values (T=0.12, F=0.14, S=0.22, P=0.20, H=0.18, A=0.14). \\
\addlinespace
Reference [2] & Corrected from \enquote{AI and its new challenges for human rights} to \enquote{AI4People: An Ethical Framework for a Good AI Society} (Floridi et al., 2018). \\
\addlinespace
References & Expanded from 13 to 27 verified entries, adding foundational works on stakeholder theory, MCDM, and AI governance. \\
\addlinespace
LaTeX Source & Created complete LaTeX source from scratch (no prior source existed in the repository). \\
\addlinespace
Screenshots & Captured real screenshots from localhost using Playwright, replacing placeholder text descriptions. \\
\bottomrule
\end{tabularx}
\end{table}
