\chapter{Platform Interface}
\label{ch:interface}

The MEDF platform is equipped with an interactive web-based dashboard built using Streamlit. This interface provides a user-friendly environment for analysts, developers, and other stakeholders to conduct ethical evaluations without needing to interact directly with the backend API or the codebase. The dashboard is designed to guide the user through the evaluation workflow, from configuration to results analysis. This chapter describes the critical features and components of the user interface.

\section{Main Evaluation View}
\label{sec:main_view}

The primary view of the dashboard allows the user to configure and run a new evaluation. The main panel is divided into several configurable sections:

\begin{itemize}[leftmargin=2em]
  \item \textbf{AI System Selection.} A dropdown menu to select the AI system to be evaluated, populated from the case studies available in the repository.

  \item \textbf{Baseline Dimension Scores.} A set of six sliders, one for each Unified Dimension, allows the user to input the baseline ethical performance scores of the selected AI system on a 1--7 Likert scale.

  \item \textbf{Framework and Stakeholder Selection.} Checkboxes to select the governance frameworks (ALTAI, NIST RMF, MGAF) and stakeholder profiles (Developer, Regulator, Affected Community) to include in the analysis.

  \item \textbf{Stakeholder Weight Customization.} An expandable section where the user can view and override the default priority weights for each selected stakeholder, providing complete control over the evaluation context.
\end{itemize}

Once the configuration is set, the user can click the \enquote{Run Evaluation} button. The dashboard then sends a request to the backend API and displays the results in real-time. The interface also provides a \enquote{Run Demo Scenario} button that populates the form with pre-configured case study data for quick demonstration and testing.

\section{Results Visualization}
\label{sec:results_viz}

After an evaluation is complete, the results are presented in a series of clear and informative visualizations:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Overall Score Gauge.} A prominent gauge visual displays the final, aggregated ethical score of the AI system, with color-coding to indicate the corresponding risk level (e.g., green for Low, red for Critical).

  \item \textbf{Framework Score Breakdown.} A bar chart compares the system's performance under each selected governance framework, allowing the user to see if the system performs better against certain standards than others.

  \item \textbf{Dimensional Contribution Radar Chart.} A radar chart visualizes the contribution of each of the six Unified Dimensions to the final score. This helps the user quickly identify the ethical strengths and weaknesses of the system.
\end{itemize}

\section{Conflict and Consensus Analysis}
\label{sec:conflict_viz}

The dashboard also includes dedicated views for analyzing stakeholder disagreement:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Conflict Matrix.} A heatmap displays the Spearman's rank correlation coefficient for each pair of stakeholders, providing an at-a-glance overview of where the most significant conflicts in priorities lie.

  \item \textbf{Pareto Frontier Visualization.} When the Pareto optimization is run, the interface displays a plot of the resulting non-dominated consensus solutions. This allows the user to visually explore the trade-offs between different compromise solutions.
\end{itemize}

\section{Case Study Browser}
\label{sec:case_browser}

The dashboard includes a dedicated \enquote{Case Studies} tab that allows the user to browse and run the three pre-configured case studies (Facial Recognition, Hiring Algorithm, Healthcare Diagnostic). Each case study entry displays a summary of the AI system, its deployment context, known risks, and the evidence sources used to derive the baseline scores. The user can select a case study and run a complete evaluation with a single click, making it easy to reproduce the results presented in Chapter~\ref{ch:case_studies}.

By providing these interactive visualizations, the Streamlit dashboard makes the complex, multi-dimensional results of the MEDF evaluation accessible and interpretable for a non-technical audience, facilitating a more informed and data-driven discussion about AI ethics.
